---
title: "An introduction to odmind"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to odmind}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
This package consists of several functions which allow you to calculate 
origin-destination matrices based on the *OpenStreetMap* road network using the
[openrouteservice package](https://github.com/GIScience/openrouteservice-r). 
The core functions build upon that and allow you to **calculate accessibility 
indicators** and **aggregate indicators** to spatial areas of interest.

```{r setup}
# load the package
library(odmind)
```
********
## Step 0 - Explore included data *(optional)*
For visualization purposes a dataset with precalculated results
and a `sf` of geometry type `POLYGON` is provided.

### testdata - odm-object (list)
```{r echo=TRUE, message=FALSE}
# Load the package data
data("testdata")
# Basic information about the odm-list object
knitr::kable(summary(testdata$od_result))
```
### testdata - spatial data
```{r echo=TRUE, message=FALSE, fig.align='center', fig.width= 6.5}
library(sf)
library(dplyr)
# A preview of the data included in the package
# create pnt_sf using the included x, y coordinates, EPSG:4326
# needs to be projected to EPSG:3035 for plotting
targets_sf <- sf::st_as_sf(testdata$od_result$targets, coords = c("x", "y")) %>%
  st_set_crs(4326) %>%
  st_transform(3035)
par(mar = c(0,0,1,0))
# 500m polygon grid for aggregation purposes, EPSG:3035 
plot(testdata$grid_500m$geometry, border = "darkgrey", reset = FALSE)
# source point data, n = 200, EPSG:4326
# needs to be projected to EPSG:3035 for plotting
plot(st_transform(testdata$od_result$sources, 3035)$geom, add = TRUE)
# target point data, n = 163
plot(targets_sf$geometry, col = "red", pch = 20, cex = .8, add = TRUE)
```

## Step 1 - Create an odm-object
All functions with the suffix `create_` work with an odm-object which is 
of type `list` and contains a `numeric matrix` for *distance* and *duration* as 
well as the input data that was used to calculate the matrices. The sources are 
a `sf` of geometry type `POINT` and the targets a `data.frame`. 

To make full use of all package functions the sources and targets need to 
follow at least the following data structures. Additional columns with further 
attributes can be included in the sources and targets if needed. Coordinates 
need to be supplied in `EPSG:4326`.

```{r}
knitr::kable(head(testdata$od_result$sources, 5))
knitr::kable(head(testdata$od_result$targets, 5))
```

### Execute calculaton

An odm-object can be created sequentially by using
```{r eval=FALSE, include=TRUE}
options(openrouteservice.url = "PUT YOUR URL HERE")
odm_result <- calc_odm(profile_string = "foot-walking",
                       source_locations = testdata$od_result$sources,
                       target_locations = testdata$od_result$targets,
                       max_chunk_size = 200)
```

or in parallel by using

```{r eval=FALSE, include=TRUE}
handlers(global = TRUE)
handlers("progress")
registerDoFuture()
plan(multisession, gc = TRUE)
res <- calc_odm_multicore(profile_string = "foot-walking",
                          sources = sources,
                          targets = targets,
                          max_chunk_size = 200,
                          ors_url = "PUT YOUR URL HERE")
```

## Step 2 - Calculate accessibility indicators
Once you have created a odm-list-object calculating indicators is straight 
forward. The following functions are provided:

| function | arguments | short description | output |
| :------- | :-------- | :---------------- | :-----: |
| `create_time_distance_sf()` | `odm-object` | Calculates distance in meters and duration in minutes from all source points to the closest target location | `sf` geometry type `POINT` |
| `create_time_distance_n_sf()` | `odm-object`, n | Calculates distance in meters and duration in minutes from all source points to n closest target locations | `sf` geometry type `POINT` |
| `create_time_distance_by_cat_sf()` | `odm-object`, filter_attribute, filter_value | Calculates distance in meters and duration in minutes from all source points to the closest target location of a chosen category | `sf` geometry type `POINT` |
| `create_time_distance_n_by_cat_sf()` | `odm-object`, filter_attribute, filter_value, n | Calculates distance in meters and duration in minutes from all source points to n closest target locations of a chosen category | `sf` geometry type `POINT` |
| `create_cumulative_sf()` | `odm-object`, filter_value_type, accessibility_filter_value, search_direction | All locations which are accessible within the applied time or distance threshold are counted. The count can be calculated in either direction (source to target or target to source) | `sf` geometry type `POINT` |
| `create_cumulative_by_cat_sf()` | `odm-object`, filter_value_type, accessibility_filter_value, search_direction, filter_attribute, filter_value | All locations which belong to a chosen category and are accessible within the applied time or distance threshold are counted. The count can be calculated in both directions (source to target or target to source) | `sf` geometry type `POINT` |
| `create_potential_sf()` | `odm-object`, filter_value_type, accessibility_filter_value, search_direction, ors_profile | A decay function is applied to all locations which are accessible within the defined time or distance threshold. The function models a location potential respecting the spatial distribution. Thus, closer features result in higher values. It can be calculated in either direction (source to target or target to source) | `sf` geometry type `POINT` |
| `create_potential_by_cat_sf()` | `odm-object`, filter_value_type, accessibility_filter_value, search_direction, ors_profile, filter_attribute, filter_value | A decay function is applied to all locations which belong to a chosen category and are accessible within the defined time or distance threshold. The function models a location potential respecting the spatial distribution. Thus, closer features result in higher values. It can be calculated in either direction (source to target or target to source) | `sf` geometry type `POINT` |

The following example shows the calculation of the distance and duration to the nearest target location using the provided dataset:
```{r message=FALSE, fig.align='center', fig.width= 6.5}
result <- create_time_distance_sf(testdata$od_result)
par(mar = c(0,0,1,0))
# Min_Dist could be switched to Min_Time
plot(result["Min_Dist"], key.pos = 1)
```

********
This is only the most basic example but the other listed functions work in a
similar way. You just need to add further function parameters.

## Step 3 - Aggregate an indicator to an area of interest
Once you have calculated one or more **accessibility indicators** you might want 
to aggregate those to statistical spatial units or regular grids. For this use 
case corresponding aggregation functions are provided:

| function | arguments | short description | output |
| :------- | :-------- | :---------------- | :-----: |
| `aggregate_expenses()` | aoi_sf, pnt_sf, id_col, crs | Aggregate the `sf` (geometry type `POINT`) attributes *Min_Dist* and *Min_Time* to an area of interest | `sf` geometry type `POINT` |
| `aggregate_accumulation()` | aoi_sf, pnt_sf, id_col, crs | Aggregate the `sf` (geometry type `POINT`) attribute *MeanCnt* or *MedianCnt* to an area of interest | `sf` geometry type `POINT` |
| `aggregate_potential()` | aoi_sf, pnt_sf, id_col, crs | Aggregate the `sf` (geometry type `POINT`) attributes *MeanPot* and *MedianPot* to an area of interest | `sf` geometry type `POINT` |
| `aggregate_coverage_rate_by_threshold()` | aoi_sf, pnt_sf, id_col, crs | Calculate an average coverage rate within an area of interest by applying a threshold filter to one of the `sf` (geometry type `POINT`) attributes *Min_Dist* or *Min_Time*. A source location is defined as covered if a target location is accessible within the threshold distance or time. | `sf` geometry type `POINT` |

The following example shows the aggregation and visualization of the minimum 
distance and duration values calculated in *step 2*:
```{r message=FALSE, fig.align='center', fig.width=6.5}
library(dplyr)
min_dist_time_sf <- create_time_distance_sf(testdata$od_result) %>%
  sf::st_transform(4647)
# Add targets for plotting
targets_sf <- sf::st_as_sf(testdata$od_result$targets, coords = c("x", "y")) %>%
  sf::st_set_crs(4326) %>%
  sf::st_transform(4647)
result_sf <- aggregate_expenses(aoi_sf = testdata$grid_500m,
                                pnt_sf = min_dist_time_sf,
                                id_col = "id",
                                crs = 4647)
par(mar = c(0,0,1,0))
# MedianDist could be switched to MedianTime
plot(result_sf["MedianDist"], border = "darkgrey", key.pos = 1, reset = FALSE)
plot(min_dist_time_sf$geom, pch = 20, cex = 1, add = TRUE)
plot(targets_sf$geometry, col = "red", pch = 20, cex = .8, add = TRUE)
```

********
This is only one example but the other listed functions work in a similar way. 
You just need to provide the respective accessibility indicator result to the 
aggregation function.

## Step 4 - Merge multiple aggregated indicator results *(optional)*
For convenience a merge function is provided which allows you to easily combine
the `sf` objects of multiple **aggregated indicators** which have been 
aggregated to the same areas of interest.

```{r message=FALSE}
# Travel expenses
min_dist_time_sf <- create_time_distance_sf(testdata$od_result) 
expenses_sf <- aggregate_expenses(aoi_sf = testdata$grid_500m,
                                  pnt_sf = min_dist_time_sf,
                                  id_col = "id",
                                  crs = 4647)
# Accumulation 
cnt_sf <- create_cumulative_sf(odm_object = testdata$od_result,
                               filter_value_type = "distance",
                               accessibility_filter_value = 750,
                               search_direction = "to_target")
cumulative_sf <- aggregate_accumulation(aoi_sf = testdata$grid_500m,
                                        pnt_sf = cnt_sf,
                                        id_col = "id",
                                        crs = 4647)
# Potential
pot_sf <- create_potential_sf(odm_object = testdata$od_result,
                              filter_value_type = "distance",
                              accessibility_filter_value = 750,
                              search_direction = "to_target",
                              ors_profile = "foot-walking")
potential_sf <- aggregate_potential(aoi_sf = testdata$grid_500m,
                                    pnt_sf = pot_sf,
                                    id_col = "id",
                                    crs = 4647)
# Coverage rate in %
coverage_sf <- aggregate_coverage_rate_by_threshold(aoi_sf = testdata$grid_500m,
                                                    pnt_sf = min_dist_time_sf,
                                                    id_col = "id",
                                                    threshold_col = "Min_Dist",
                                                    threshold_value = 350,
                                                    crs = 4647)
# Merge 
merged_sf <- merge_indicators(aoi_sf = testdata$grid_500m, id_col = "id",
                              expenses_sf, cumulative_sf, potential_sf,
                              coverage_sf)
# Remove the the geometry column just for table presentation
knitr::kable(head(sf::st_drop_geometry(merged_sf), 5))
```

This last example gives a concluding first impression of the functions provided
within package. See `?odmind` for more details!

